Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers

### 2019-09-10 Manuscript/manuscriptPS.tex
Counterion binding

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between the lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers compared to PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups.
We explored this by quantifying the counterion density profiles along the membrane normal, accompanied by the areas per lipid, see Fig.~\ref{NAdensPOPS}.
The force fields studied show significant differences in both binding affinity
and distribution of ions at the interface.
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
is reproduced only in GROMOS-CKP and in the MacRog simulation
with potassium counterions, while other models give considerably smaller values (Fig.~\ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per molecule values
since the MacRog simulation, which has the strongest sodium binding
(the lowest concentrations in bulk water), gives the same area per molecule
as the CHARMM36-UA simulation, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulations with potassium produce a larger area per molecule (63~\AA$^2$) than
with sodium (53~\AA$^2$) in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
The results are consistent with a previous study suggesting that the
low areas per molecule in PS lipid bilayers originate from the combination
of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.

\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPSformatted.pdf}
  \caption{\label{NAdensPOPS}
    Counterion density profiles along the membrane normal and areas per lipid (A$_{\rm pl}$)
    of POPS lipid bilayer from simulations with different force fields.
    The vertical green bars indicate the location of the phosphate density peak.
    The experimental area per lipid is~62.7~$\pm$~1.3~\AA$^2$~\cite{pan14}.
  }
\end{figure}

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
    Experimental values are from Ref. \citenum{scherer87}, with the signs measured in Ref.~\citenum{ferreira16}.
    The values from the CHARMM36 and GROMOS-CKP simulations with sodium are averages of two independent simulations and
    the error bar is given as the difference between the results divided by two.
  }\end{figure}

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
  \caption{\label{HGorderparametersPSvsPC}
    Modulation of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
    Experimental values with the signs are measured for pure POPS system in this work.
    The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
    Because the experimental values of POPS in pure and mixed bilayers come from $^{13}$C NMR (this work) and $^2$H NMR (Ref. \citenum{roux90}), respectively,
    the error bars of 0.02 are used here \cite{botan15,ollila16}.
    The y-axis for the $\alpha$-carbon results of POPS (bottom) is shifted
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
    The values from CHARMM36 and GROMOS simulations are averages of two independent simulations and
    the error bars are given as the difference between the results divided by two. 
  }
\end{figure}

The experimentally observed modulation of headgroup order parameters
by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate the cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
Studying binding of cations to negatively charged lipid bilayers is less straighforward due to the presence of cationic
counterions (the lack of an ion-free reference state). The analysis is further complicated by the artificial aggregation of counterions
observed in some simulations (section \ref{mixtureTOadditionalCIs} in the SI).
Therefore, we evaluate here the amount of bound charge not by adding salt
(although this is discussed in the SI section \ref{mixtureTOadditionalCIs}),
but by studying the changes of the headgroup $S_{\rm CH}$ order parameters with increasing amount of
negatively charged lipids (and thus increasing amount of cationic counterions) in the bilayer.

Experimentally, the $S_{\rm{CH}}$ values of the $\alpha$ and $\beta$ headgroup carbons of POPC increase when negatively charged POPS lipids are incorporated in the bilayer 
(section \ref{HGorderparametersPCvsPEPSPGchol})~\cite{seelig87,scherer87}.
This is reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
which have the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
The CHARMM36, Berger and GROMOS-CKP simulations either exhibit no change or show a decrease
in either or both the POPC headgroup order parameters as the amount of POPS increases (Fig. \ref{HGorderparametersPCvsPS}).
Therein, the stronger counterion binding cancels
the effect of negatively charged headgroups and prevents the experimentally observed
increase of headgroup order parameters with growing amount of PS lipids.
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces the most
realistic surface charge density in membranes containing PS lipids,
while the other tested models overestimate the counterion
binding affinity. The results are consistent with the behavior of headgroup order
parameters as a function of added counterions analyzed in section \ref{mixtureTOadditionalCIs}
in the SI.

The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
together with other experimental results suggest that the PS headgroup structure becomes less rigid when diluted with
POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
Unfortunately, none of the tested models correctly reproduce the modulation of POPS headgroup order
parameters with increasing amount of POPC in POPC:POPS mixtures (Fig. \ref{HGorderparametersPSvsPC}).
More accurate force fields are needed
to correctly describe the PC--PS headgroup interactions in MD simulations.

>>>>

Membranes containing PS lipids are always accompanied with counterions that
modulate electrostatic interactions between the lipids and other biomolecules. MD simulations have suggested
that counterions reduce the area per lipid of PS bilayers compared to PC
bilayers~\cite{pandit02,mukhopadhyay04,pedersen06} by screening the repulsion between charged lipid headgroups.
We explored this by quantifying the counterion density profiles along the membrane normal, accompanied by the areas per lipid, $\langle A_\mathrm{pl} \rangle$, see Fig.~\ref{NAdensPOPS}.
The force fields studied showed significant differences in both the binding affinity
and the distribution of ions at the interface.
The experimental area per lipid (62.7~$\pm$~1.3~\AA$^2$) \cite{pan14} 
was reproduced only in GROMOS-CKP and GROMOS-CKPM, as well as in MacRog
with potassium counterions, while all other models gave considerably smaller values (Fig.~\ref{NAdensPOPS}).
However, the counterion binding and the concomitant electrostatic screening of the headgroup
repulsion does not fully explain the low area per lipid values
since MacRog, which has the strongest sodium binding
(the lowest concentrations in bulk water), gave the same $\langle A_\mathrm{pl} \rangle$
as CHARMM36-UA, which has significantly weaker counterion binding
affinity. On the other hand, MacRog simulation with potassium produced a larger area per lipid (63~\AA$^2$) than
with sodium (53~\AA$^2$), in line with the weaker potassium binding affinity (Fig.~\ref{NAdensPOPS}).
The results are consistent with a previous study that suggested the
low $\langle A_\mathrm{pl} \rangle$ in PS lipid bilayers to originate from the combination
of both the counterion binding and intermolecular interactions between lipid headgroups~\cite{petrache04}.

\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/NAdensPOPSformatted.pdf}
  \caption{\label{NAdensPOPS}
    Counterion density profiles along the membrane normal and areas per lipid, $\langle A_{\rm pl} \rangle$,
    of POPS lipid bilayer from simulations with different force fields.
    The vertical bars indicate the locations of the phosphate density peaks.
    The experimental area per lipid is~62.7~$\pm$~1.3~\AA$^2$~\cite{pan14}.
  }
\end{figure}

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
    Experimental values are from Ref. \citenum{scherer87}, with the signs measured in Ref.~\citenum{ferreira16}.
    The values from the CHARMM36 and GROMOS-CKP simulations with sodium are averages of two independent simulations,
    the error bars show the difference between the results divided by two.
  }\end{figure}

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
  \caption{\label{HGorderparametersPSvsPC}
    Modulation of the POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
    Experimental values with the signs were measured for a pure POPS system in this work.
    The signs are assumed to remain unchanged for the mixture, and the magnitudes are taken from Ref. \citenum{roux90}.
    Because the experimental values of POPS in pure and mixed bilayers come from $^{13}$C NMR (this work) and $^2$H NMR (Ref. \citenum{roux90}), respectively,
    the error bars of 0.02 are used here \cite{botan15,ollila16}.
    The $y$-axis for the $\alpha$-carbon results (bottom) is shifted
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
    The values from CHARMM36 and GROMOS simulations are averages of two independent simulations,
    the error bars show the difference between the results divided by two. 
  }
\end{figure}

The experimentally observed modulation of headgroup order parameters
by increasing salt concentration (the molecular electrometer concept) has been previously used to evaluate cation binding to zwitterionic PC bilayers in simulations \cite{catte16}.
Studying the binding of cations to negatively charged lipid bilayers is less straightforward, because an ion-free reference state does not exist due to the ever-present cationic
counterions. Our analysis was further complicated by the artificial aggregation of counterions
observed in some simulations (section \ref{mixtureTOadditionalCIs} in the Supporting Information).
Therefore, we evaluate here the amount of bound charge not by adding salt
(although this is discussed in section \ref{mixtureTOadditionalCIs}),
but by studying the changes of the headgroup $S_{\rm CH}$ with an increasing amount of
negatively charged lipids (and thus an increasing amount of cationic counterions) in the bilayer.

Experimentally, the $S_{\rm{CH}}$ values of the $\alpha$ and $\beta$ headgroup carbons of POPC increase when negatively charged POPS lipids are incorporated in the bilayer 
(section \ref{electrometerFORmixtures})~\cite{seelig87,scherer87}.
This was reproduced in the MacRog simulations with potassium counterions (Fig. \ref{HGorderparametersPCvsPS}),
which had the weakest binding affinity to POPS lipid bilayers (Fig.~\ref{NAdensPOPS}).
The CHARMM36, Berger, and GROMOS-CKP simulations either exhibited no change, or showed a decrease
in one or even both of the POPC headgroup order parameters, as the amount of POPS increased (Fig. \ref{HGorderparametersPCvsPS}).
Therein, the stronger counterion binding canceled
the effect of negatively charged headgroups and prevented the experimentally observed
increase of headgroup order parameters with growing amount of PS lipids.
Therefore, we suggest that the relatively weak binding of potassium
in the MacRog simulations (Fig. \ref{NAdensPOPS}) produces a more
realistic surface charge density in membranes containing PS lipids
than the other tested models that overestimate the counterion
binding affinity. The results are consistent with the behavior of headgroup order
parameters as a function of added counterions, see section \ref{mixtureTOadditionalCIs}
in the Supporting Information.

The reduced forking of the POPS $\alpha$-carbon (Fig.~\ref{HGorderparametersPSvsPC})
together with other experimental results suggest that the PS headgroup structure becomes less rigid when diluted with
POPC~\cite{browning80,buldt81,roux90,roux91,scherer87}.
Unfortunately, none of the tested models correctly reproduced the modulation of POPS headgroup order
parameters with increasing amount of POPC %in POPC:POPS mixtures
(Fig. \ref{HGorderparametersPSvsPC}).
More accurate force fields are needed
to correctly describe the PC--PS headgroup interactions in MD simulations.
